# Data

## Description

This data comes from the United States Department of Transportation’s Bureau of Transportation Statistics and contains information on flight delays and cancellations for domestic flights in the first six months of 2025. In that time period, there were a total of 11,404 delayed operations by 21 airlines at 370 airports.

Each observation represents aggregated monthly statistics for a specific airport-airline combination (i.e. HOU-Southwest-January or IAD-Delta-May). There are 21 variables representing flight volumes, delay counts, delay durations, cancellations and diversions. A flight is considered delayed if it arrives more than 15 minutes after its scheduled time.

Key Delay Categories (each measured in both count and minutes): 

-   Carrier delays: airline-controlled issues (maintenance, crew, baggage) 

-   Weather delays: extreme weather conditions

-   NAS delays: National Airspace System issues (air traffic control, non-extreme weather, high volume) 

-   Security delays 

-   Late aircraft delays 

\
It is important to note that the delay time and delay count metrics capture different aspects of performance; an airport could have many short delays (high count, low time) or a few very long delays (low count, high time), so both dimensions must be considered in the analysis. Additionally, because airports and airlines vary significantly in size, performance comparisons will need to be normalised by the total number of incoming flights to create meaningful per-flight delay rates. 

## Missing value analysis

```{r}
#| eval: false
df <- read_csv("Airline_Delay_Cause.csv")
colSums(is.na(df)) |> 
  sort(decreasing = TRUE)
```

```{r}
missing_percent <- df %>%
  summarise(across(everything(), ~ mean(is.na(.))*100)) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "percent_missing")

ggplot(missing_percent, aes(x = reorder(variable, percent_missing), y = percent_missing)) +
  geom_col(fill = "blue") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Percent Missing per Variable",
    y = "% Missing"
  )

```

The missing data rate is extremely small; the highest missing varaible rate of all column is around 0.05%. Overall, this dataset contains a negligible amount of missing data, so we are not taking any further process to handle missing data.
