# Results

## Q1. How do delay patterns vary across U.S. airports and states, and how are these patterns associated with geography, dominant delay types, and airport traffic volume?

```{r}
library(tidyverse)
library(GGally)
library(janitor)
library(statebins)
library(ggplot2)

df <- read_csv("Airline_Delay_Cause.csv")
```

```{r}
delay_airport <- df |>
  group_by(airport, airport_name) |>
  summarise(
    total_flights = sum(arr_flights, na.rm = TRUE),
    carrier_delay = sum(carrier_delay, na.rm = TRUE),
    weather_delay = sum(weather_delay, na.rm = TRUE),
    nas_delay = sum(nas_delay, na.rm = TRUE),
    security_delay = sum(security_delay, na.rm = TRUE),
    late_aircraft_delay = sum(late_aircraft_delay, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    carrier_rate = carrier_delay / total_flights,
    weather_rate = weather_delay / total_flights,
    nas_rate = nas_delay / total_flights,
    security_rate = security_delay / total_flights,
    late_aircraft_rate = late_aircraft_delay / total_flights
  )

delay_cols <- c("carrier_rate", "weather_rate", "nas_rate",
                "security_rate", "late_aircraft_rate")

delay_norm <- delay_airport |>
  mutate(across(all_of(delay_cols), scale))

library(viridis)

delay_cols <- c(
  "carrier_rate",
  "weather_rate",
  "nas_rate",
  "security_rate",
  "late_aircraft_rate"
)
```

```{r}
library(tidyverse)
library(GGally)

state_delay <- delay_airport |>
  mutate(
    state = sub(".*, ([A-Z]{2}):.*", "\\1", airport_name),
    region = state.region[match(state, state.abb)]
  ) |>
  group_by(state, region) |>
  summarise(
    carrier_rate = mean(carrier_rate, na.rm = TRUE),
    weather_rate = mean(weather_rate, na.rm = TRUE),
    nas_rate = mean(nas_rate, na.rm = TRUE),
    security_rate = mean(security_rate, na.rm = TRUE),
    late_rate = mean(late_aircraft_rate, na.rm = TRUE),
    .groups = "drop"
  )

GGally::ggparcoord(
  data = state_delay,
  columns = 3:7,        # the 5 delay rate columns
  groupColumn = "region",
  scale = "std",        # normalize each axis
  alphaLines = 0.6
) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Parallel Coordinate Plot of State Delay Profiles",
    subtitle = "Lines represent states; Colors represent Census regions",
    x = "Delay Type",
    y = "Standardized Delay Rate",
    color = "Region"
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

-   Western states tend to have lower rates cross all delay types, and security delay rate is the smallest, which is shown as distinct cluster around 0.

```{r}
library(tidyverse)
library(reshape2)
library(viridis)
library(ggplot2)

delay_cols <- c(
  "carrier_rate",
  "weather_rate",
  "nas_rate",
  "security_rate",
  "late_aircraft_rate"
)

corr_mat <- delay_airport |>
  select(all_of(delay_cols)) |>
  cor(use = "pairwise.complete.obs")

corr_df <- melt(corr_mat)
names(corr_df) <- c("DelayType1", "DelayType2", "Correlation")

ggplot(corr_df, aes(x = DelayType1, y = DelayType2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_distiller(
  palette = "Blues", direction=1) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Correlation Heatmap of Airport Delay Types",
    x = "Delay Type",
    y = "Delay Type",
    fill = "Correlation"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

-   Security delays shows almost no correlations with other delay type. Weather and carrier rate, and late aircraft and NAS rate are most correlated.

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

delay_scatter_long <- delay_airport |>
  select(
    airport, total_flights,
    carrier_rate, weather_rate, nas_rate,
    security_rate, late_aircraft_rate
  ) |>
  pivot_longer(
    cols = ends_with("_rate"),
    names_to = "delay_type",
    values_to = "delay_rate"
  )

ggplot(delay_scatter_long,
       aes(x = total_flights, y = delay_rate)) +
  geom_point(alpha = 0.5, size = 1.5, color = "blue") +
  scale_x_log10(
    breaks = c(100, 1000, 10000, 100000),
    labels = c("100", "1K", "10K", "100K")
  ) +
  facet_wrap(~ delay_type, scales = "free_y") +
  theme_minimal(base_size = 15) +
  labs(
    title = "Airport Size vs Delay Rate by Delay Type",
    x = "Total Flights (Log Scale)",
    y = "Delay Rate (%)"
  )
```

-   Airport size does not strongly predict overall delay rate. The delay rate varies more in the airports with smaller airport size (around 1k), often including extreme outliers.

```{r, echo=FALSE, results='hide'}
delay_airport <- delay_airport |>
  mutate(
    state = sub(".*, ([A-Z]{2}):.*", "\\1", airport_name)
  )

delay_airport |> 
  select(airport, airport_name, state) |> 
  head(10)

state_summary <- delay_airport |>
  mutate(state = sub(".*, ([A-Z]{2}):.*", "\\1", airport_name)) |>
  group_by(state) |>
  summarise(
    avg_delay_rate = mean((carrier_rate + weather_rate + nas_rate +
                           security_rate + late_aircraft_rate) / 5, na.rm = TRUE),
    total_flights = sum(total_flights, na.rm = TRUE)
  )
```

```{r}
library(ggalluvial)

state_flow <- delay_airport |>
  mutate(
    state = sub(".*, ([A-Z]{2}):.*", "\\1", airport_name),
    region = state.region[match(state, state.abb)],
    dominant = delay_cols[max.col(select(., all_of(delay_cols)), 
                                  ties.method = "first")]
  ) |>
  drop_na(region) |>      
  group_by(region, dominant) |>
  summarise(count = n(), .groups = "drop")

ggplot(state_flow,
       aes(axis1 = region, axis2 = dominant, y = count)) +
  geom_alluvium(aes(fill = dominant)) +
  geom_stratum() +
  geom_label(aes(label = after_stat(stratum)), stat = "stratum") +
  theme_minimal() +
  labs(
    title = "Region-to-Dominant Delay Type Flow",
    y = "Number of Airports"
  )

```

-   Late aircraft delays are more dominant type for most airports nationwide. Carrier delays dominate many airports in the South and West.

```{r}
library(sf)
library(tidyverse)

delay_by_state_type <- delay_airport |>
  mutate(state = sub(".*, ([A-Z]{2}):.*", "\\1", airport_name)) |>
  group_by(state) |>
  summarise(
    carrier_delay = sum(carrier_delay, na.rm = TRUE),
    weather_delay = sum(weather_delay, na.rm = TRUE),
    nas_delay = sum(nas_delay, na.rm = TRUE),
    security_delay = sum(security_delay, na.rm = TRUE),
    late_aircraft_delay = sum(late_aircraft_delay, na.rm = TRUE),
    total_flights = sum(total_flights, na.rm = TRUE)
  ) |>
  mutate(
    state_full = tolower(state.name[match(state, state.abb)]),
    carrier_rate = carrier_delay / total_flights,
    weather_rate = weather_delay / total_flights,
    nas_rate = nas_delay / total_flights,
    security_rate = security_delay / total_flights,
    late_rate = late_aircraft_delay / total_flights
  )

```

```{r}
library(ggplot2)
library(viridis)

delay_long <- delay_by_state_type |>
  pivot_longer(
    cols = ends_with("_rate"),
    names_to = "delay_type",
    values_to = "delay_value"
  ) |>
  mutate(
    delay_type = recode(delay_type,
                        carrier_rate = "Carrier",
                        weather_rate = "Weather",
                        nas_rate = "NAS",
                        security_rate = "Security",
                        late_rate = "Late Aircraft")
  )

states_sf <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))

plot_data <- left_join(
  states_sf,
  delay_long,
  by = c("ID" = "state_full")
) |>
  mutate(
    state_abb = toupper(state.abb[match(ID, tolower(state.name))]),
    region = state.region[match(state_abb, state.abb)]  
  ) |>
  filter(!is.na(region)) 

delay_types <- plot_data |>
  dplyr::filter(!is.na(delay_type)) |>
  dplyr::pull(delay_type) |>
  unique()

plot_list <- list()

for (dt in delay_types) {
  p <- ggplot(filter(plot_data, delay_type == dt)) +
    geom_sf(aes(fill = delay_value), color = "white", linewidth = 0.2) +
    geom_sf_text(aes(label = state_abb), size = 3, color = "black") +
    scale_fill_viridis_c(direction = -1) +
    theme_void() +
    labs(
      title = paste(dt, "Delay Rate per Flight by State"),
      fill = "Delay per Flight"
    )

  plot_list[[dt]] <- p
}

plot_list

```

-   Carrier

    -   There are unusual high delays in West Virginia, and parts of Mid-Atlantic states. Most of other states have moderate and consistent delays.

-   Weather

    -   The delay rates are higher in Midwest and Mountain states. Again, West Virginia has the most extreme delay rate, which aligns with previous delay type.

-   NAS delays

    -   Overall delays are relatively low. Colorado has relatively high NAS delays, and New Jersey stands out as an outline with very high delay rate.

-   Security

    -   Security delays are extremely rare in the nation, and again, West Virginia has the highest, extreme value compared to the rest of the states.

-   Late aircraft delays

    -   Delays are higher in Northeast and part of Midwest (Pennsylvania, West Virginia), while most Western and Southern states have low delay rates.
